---
title: "bb16.data.exploration"
author: "Kendra Maas"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true # table of content true
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("ggplot2")
# install.packages("vegan")
# install.packages("ecodist")
# install.packages("RColorBrewer")
# install.packages("tidyr")
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("indicspecies")
library(ggplot2)
library(vegan)
library(ecodist)
library(RColorBrewer)
library(tidyr)
library(tidyverse)
library(reshape2)
library(plyr)
library(indicspecies)
library(heatmap3)
library(devEMF)
library(dplyr)



ggplotRegression <- function (fit) {
    
    # require(ggplot2)
    
    ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
        geom_point() +
        stat_smooth(method = "lm", col = "red") +
        labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                           "Intercept =",signif(fit$coef[[1]],5 ),
                           " Slope =",signif(fit$coef[[2]], 5),
                           " P =",signif(summary(fit)$coef[2,4], 5)))
}

parseDistanceDF = function(phylip_file) {

   # Read the first line of the phylip file to find out how many sequences/samples it contains
    temp_connection = file(phylip_file, 'r')
    len = readLines(temp_connection, n=1)
    len = as.numeric(len)
    len = len +1
    close(temp_connection)
   
   
    phylip_data = read.table(phylip_file, fill=T, row.names=1, skip=1, col.names=1:len)
    colnames(phylip_data) <- row.names(phylip_data)
    return(phylip_data)
}




```

##Read in data

```{r read in data}

# otu <- read.table(file = "../bb16.oc.trim.contigs.good.unique.good.filter.precluster.pick.pick.opti_mcc.0.03.subsample.shared", header=T, stringsAsFactors = FALSE, row.names=2)
# otu <- select(otu, -label, -numOtus)
# 
# 
# taxa <- read.table(textConnection(gsub("\\(.+?\\);", "\t", readLines("../bb16.oc.trim.contigs.good.unique.good.filter.precluster.pick.pick.opti_mcc.0.03.cons.taxonomy"))), col.names=c("OTU", "Size", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus"), skip=1)
# taxa <- taxa[taxa$OTU %in% names(otu),]
# 
# # get OTU abundance for this subsampling
# sub.size <- data.frame(OTU = colnames(otu), size.sub = colSums(otu))
# 
# taxa <- full_join(taxa, sub.size, by = "OTU", copy=TRUE)
# 
# 
# maxab <- apply(otu, 2, max)
# n1 <- names(which(maxab < 10))
# otu.ab <- otu[,-which(names(otu) %in% n1)]
# taxa.ab <- taxa[-which(taxa$OTU %in% n1),]
# 
# 
# otu.total <- otu
# taxa.total <- taxa
# 
# otu <- otu.ab
# taxa <- taxa.ab
# 
# write.csv(otu, file="otu.maxab10.csv")
# write.csv(taxa, file="taxa.maxab10.csv")

#### Reading in the full OTU matrix takes a long time, storing otu and taxa with max abundance 10. then reading that in each time. Also added "preferred" to taxa

otu <- read.csv(file="otu.maxab10.csv", header=T, stringsAsFactors = FALSE, row.names=1)
taxa <- read.csv(file="taxa.maxab10.csv")
taxa$preferred <- factor(taxa$preferred, levels=taxa$preferred[order(taxa$Phylum)], ordered = TRUE)
taxa$preferred <- factor(taxa$preferred[,drop=TRUE])

# get OTU abundance for this subsampling

maxab <- apply(otu, 2, max)
n1 <- names(which(maxab < 200))
otu.ab <- otu[,-which(names(otu) %in% n1)]
taxa.ab <- taxa[-which(taxa$OTU %in% n1),]

# Alpha diversity

alpha <- read.table(file="../bb16.oc.trim.contigs.good.unique.good.filter.precluster.pick.pick.opti_mcc.groups.ave-std.summary", header=T, stringsAsFactors = FALSE)
alpha <- filter(alpha, label== "0.03" & method == "ave")

# Beta diversity
jc <- parseDistanceDF("../bb16.oc.trim.contigs.good.unique.good.filter.precluster.pick.pick.opti_mcc.jest.0.03.lt.ave.dist")
bc <- parseDistanceDF("../bb16.oc.trim.contigs.good.unique.good.filter.precluster.pick.pick.opti_mcc.braycurtis.0.03.lt.ave.dist")
tyc <- parseDistanceDF("../bb16.oc.trim.contigs.good.unique.good.filter.precluster.pick.pick.opti_mcc.thetayc.0.03.lt.ave.dist")

# experimental data
expdata <- read.table(file="../BioBlitz2016_env.txt", header=T, stringsAsFactors = FALSE, fill=TRUE)
expdata <- arrange(expdata, group)

expdata$detail.type <- factor(expdata$detail.type, levels = expdata$detail.type[order(expdata$bioblitz.transect)], ordered=TRUE)

expdata$detail.type <- factor(expdata$detail.type[,drop=TRUE])

alpha.expdata <- left_join(alpha, expdata, on="group")

bb <- alpha.expdata$experiment == "bioblitz.transect"

alpha.expdata <- alpha.expdata[bb,]
alpha.expdata$detail.type <- factor(alpha.expdata$detail.type[,drop=TRUE])
jc <- jc[bb,bb]
bc <- bc[bb,bb]
tyc <- tyc[bb,bb]
otu <- otu[bb,]
otu.ab <- otu.ab[bb,]



```

##Alpha Diversity
Boxplots of alpha diversity measures (diversity within a sample)

I'm adding the actual data points to the first one just to demonstrate the possiblility. Sometimes I like it in addition to the boxplot

```{r alpha diversity treatment}
# alpha diversity (diversity within a sample)

   


xlab <- paste(levels(as.factor(alpha.expdata$detail.type)),"\n(N=", table(alpha.expdata$detail.type),")", sep="")
ggplot(data=alpha.expdata, (aes(x=detail.type, y=sobs, fill=detail.type)))+
    geom_boxplot()+
    scale_fill_manual(values=treat.col, name= "Transect Location")+
    geom_jitter(width=0.2)+
    # scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial Richness by detail.type")+
    labs(y="Species Observed", x="")+
    facet_grid(.~alpha.expdata$urban)+
    theme(axis.text.x = element_text(angle =  45, hjust = 1))
ggsave(file="richness.pdf", width=8, height=4)

xlab <- paste(levels(as.factor(alpha.expdata$detail.type)),"\n(N=", table(alpha.expdata$detail.type),")", sep="")
ggplot(data=alpha.expdata, (aes(x=detail.type, y=shannon)))+
    geom_boxplot(varwidth = T)+
    geom_jitter(width=0.2)+
    scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial Diversity by detail.type")+
    labs(y="Shannon Diversity Index", x="")+
    facet_grid(.~alpha.expdata$urban)


xlab <- paste(levels(as.factor(alpha.expdata$detail.type)),"\n(N=", table(alpha.expdata$detail.type),")", sep="")
ggplot(data=alpha.expdata, (aes(x=detail.type, y=invsimpson, fill=detail.type)))+
    geom_boxplot()+
    scale_fill_manual(values=treat.col, name= "Transect Location")+
    geom_jitter(width=0.2)+
    # scale_x_discrete(labels=xlab)+
    theme_bw()+
    ggtitle("Bacterial Diversity by transect location")+
    labs(y="Inverse Simpson Diversity Index", x="")+
    facet_grid(.~alpha.expdata$urban)+
    theme(axis.text.x = element_text(angle =  45, hjust = 1))
ggsave(file="simp.pdf", width=8, height=4)





```

##Beta Diversity 

Non-metric multidimensional scaling to visualize between sample diversity

###jc= Jaccard dissimilarity measure presence absence
```{r beta div nms jc}

## Whole dataset


jc.nms <- metaMDS(as.dist(jc), k=2, trymin=50, trymax=500, wascores=F)
jc.points <- data.frame(jc.nms$points)
jc.plot <- ggplot(jc.points, aes(x=MDS1, y=MDS2,label=rownames(jc)))

x <- max(jc.points$MDS1)/1.5
y <- min(jc.points$MDS2)

treat.col <- c( "deep.water" = "#003c30", "marsh.water" = "#35978f", "deep.sed" = "#dfc27d",   "marsh.sed" = "#bf812d", "wet.soil" = "#8c510a", "up.soil" = "#543005")

urban.shape <- c("suburban" = 21, "urban"=24)

jc.plot + geom_point(aes(color=factor(alpha.expdata$detail.type), shape=factor(alpha.expdata$urban)), size=4)+
    theme_bw()+
    scale_color_manual(values=treat.col, name="Transect Location")+
    annotate("text", x, y, label=paste("stress = ", round(jc.nms$stress, digits = 3)))+
    labs(x="", y="")
ggsave(file="jc.detail.type.nms.jpg")





```

###bc= Bray-Curtis dissimilarity measure that incoorporates abundance
```{r beta div nms bc}

## Whole dataset


bc.nms <- metaMDS(as.dist(bc), k=2, trymin=50, trymax=500, wascores=F)
bc.points <- data.frame(bc.nms$points)
bc.plot <- ggplot(bc.points, aes(x=MDS1, y=MDS2,label=rownames(bc)))

x <- max(bc.points$MDS1)/1.5
y <- min(bc.points$MDS2)


bc.plot + geom_point(aes(fill=factor(alpha.expdata$detail.type), shape=factor(alpha.expdata$urban)), size=4)+
    theme_bw()+
    scale_fill_manual(values=treat.col, name= "Transect Location")+
    scale_shape_manual(values=urban.shape, name="Transect Location")+
    guides(fill = guide_legend(override.aes = list(shape = 22)))+
    annotate("text", x, y, label=paste("stress = ", round(bc.nms$stress, digits = 3)))+
    labs(x="", y="")
ggsave(file="bc.nms.pdf", width=8, height=6)

```


```{r beta div nms bc with vectors, warning=FALSE}
scrs <- as.data.frame(scores(bc.nms, display ='sites'))
scrs <- cbind(scrs, detail.type = alpha.expdata$detail.type, location = alpha.expdata$urban)


spec.ab <- envfit(bc.nms, otu.ab, perm=999)

spp.scrs <- as.data.frame((scores(spec.ab, display="vectors")))
spp.scrs <- cbind(spp.scrs, Species =rownames(spp.scrs))
spp.scrs <- cbind(spp.scrs, pVal=spec.ab$vectors$pvals)
# spp.scrs <- cbind(spp.scrs, Taxa=taxa$prefered, phy.col=taxa$phy.col)
spp.scrs <- cbind(spp.scrs, Phyla=taxa.ab$preferred)

###color vectors by phyla

ggplot(scrs)+
    coord_fixed()+
    geom_segment(data=spp.scrs[spp.scrs$pVal<0.0011,], 
                 aes(x=0, xend=NMDS1, y=0, yend=NMDS2, color = Phyla),
                 arrow=arrow(length=unit (0.25, "cm")), size=1.5)+
    scale_color_brewer(palette = "Spectral" )+
    geom_point(mapping=aes(x=NMDS1, y=NMDS2, fill=factor(alpha.expdata$detail.type), shape=factor(alpha.expdata$urban)), size=4)+
    theme_bw()+
    scale_fill_manual(values=treat.col, name= "Transect Location")+
    scale_shape_manual(values=urban.shape, name="Transect Location")+
    guides(fill = guide_legend(override.aes = list(shape = 22)))+
    labs(x="",y="")+
    theme_bw()
ggsave(file="bc.nms.vectors.pdf", width=8, height=7)



```

###tyc= Theta-YC dissimilarity measure that strongly weights abundance
```{r beta div nms tyc, warning=FALSE}

## Whole dataset


tyc.nms <- metaMDS(as.dist(tyc), k=2, trymin=50, trymax=500, wascores=F)
tyc.points <- data.frame(tyc.nms$points)
tyc.plot <- ggplot(tyc.points, aes(x=MDS1, y=MDS2,label=rownames(tyc)))

x <- max(tyc.points$MDS1)/1.5
y <- min(tyc.points$MDS2)


tyc.plot + geom_point(aes(fill=factor(alpha.expdata$detail.type), shape=factor(alpha.expdata$urban)), size=4)+
    theme_bw()+
    scale_fill_manual(values=treat.col, name= "Transect Location")+
    scale_shape_manual(values=urban.shape, name="Transect Location")+    annotate("text", x, y, label=paste("stress = ", round(tyc.nms$stress, digits = 3)))+
    labs(x="", y="")
ggsave(file="tyc.detail.type.nms.jpg")





```

###Hypothesis testing on beta diversity

Permanova results that I look at significance (Pr(>F)) and the R2 which can be interpreted as roughly the % varibility in your whole community that can be explained by the model factor 
```{r hypothesis testing}

permanova <- adonis(as.dist(jc)~alpha.expdata$detail.type, perm=99, rm.na=TRUE)
permanova

permanova <- adonis(as.dist(bc)~alpha.expdata$detail.type*alpha.expdata$urban, perm=99, rm.na=TRUE)

permanova

permanova <- adonis(as.dist(tyc)~alpha.expdata$detail.type*alpha.expdata$urban, perm=99, rm.na=TRUE)
permanova


```

##Indicator species 

Method for pulling out the species that coorespond with grouping variable (i.e. Treatment). See DeCaceres and Legendre
2009 for details.

```{r indicator species}

# otu[,-1] tells multipatt to ignore the first column which holds the name


indic <- multipatt(otu[,-1], as.factor(alpha.expdata$detail.type),control = how(nperm=99))

write.csv(file="indicator.species.csv",indic$sign%>%
    rownames_to_column(var = "OTU")%>%
    mutate(p.fdr = round(p.adjust(p.value, "fdr"),3))%>%
    right_join(taxa, by = "OTU")%>%
    filter(p.fdr < 0.05) %>%
    arrange(index))


sig.otus <- indic$sign%>%
    add_rownames(var="OTU")%>%
    filter(p.adjust(p.value,"fdr") < 0.05)

otu.indic <- otu[,which(names(otu) %in% sig.otus$OTU)]
taxa.indic <- taxa[which(taxa$OTU %in% sig.otus$OTU),]